---
title: "Ten Tidyverse tricks everyone should know"
description: |
  A short description of the post.
author:
  - name: Rikard Wahlström
    url: https://www.syntopicalreckoning.com/
date: 04-28-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 4
draft: TRUE
categories:
  - posts
  - tidyverse
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
theme_set(theme_minimal())

```

I recently came across a presentation by Dave Robinson on the [Lander Analytics YouTube channel](https://www.youtube.com/watch?v=NDHSBUN_rVU), in which he presents ten Tidyverse tricks. I thought I would share them here in written form, with examples for each.


Let's start by firing up the `tidyverse`!

```{r, message = FALSE, echo = TRUE}

library(tidyverse)
library(janitor)

```



**1. `Count()`**

`count()` comes from the `{dplyr}` package (which, of course, is part of the `tidyverse`). 

Dave showcases `count()` using the data from a TidyTuesday project from 2018 for recent college graduates (link [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-10-16)), so let's use the same dataset here:

```{r, message = FALSE, echo = TRUE}

tt <- tidytuesdayR::tt_load('2018-10-16')

grads <- tt$`recent-grads` %>% 
  clean_names()

```



```{r}

grads %>% 
  count(major_category,
        sort = TRUE,
        wt = total,
        name = "graduates")
```


**2. Creating variables in `count()`**

**3. `add_count()`**

**4. `summarize()` with a list column**

**5. `fct_reorder()` + `geom_col()` + `coord_flip()`**

**6. `fct_lump()`**

**7. `scale_x`/`y_log10`**

**8. `crossing()`**

**9. `separate()`**

**10. `extract()`**


